Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study

### 2020-06-03 pharmaco.Rmd
updated pharmaco manuscript

_Background_-- We previously reported that dispensing of at least one prescription from primary care in the past year was strongly associated with severe COVID-19 in Scotland, even in those without past diagnoses of conditions designated as high risk (listed conditions). Here we conducted an initial investigation of possible explanations for this association which include  specific co-morbidity or  drug effects but also include more diffuse frailty or polypharmacy effects.

@@

*Findings* -- The drug classes associated with COVID-19 included widely-prescribed cardiovascular drugs and also multiple drug classes previously identified as indicators of overprescribing without medical indications: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.  This led us to examine associations with polypharmacy, defined as the total number of drug classes dispensed.  The rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.  The association with polypharmacy was strongest in those without past hospital admissions or diagnoses of conditions previously designated as conferring high risk.  A backwards elimination procedure identified proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy.  Further analysis of this association showed a dose-response relationship with the average daily dose dispensed and that the association was strongest in younger age groups: the rate ratio for severe COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[3, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 60 years.  The association remained after adjusting for covariates prespecified as likely confounders, and in those not prescribed any of the `r nrow(coeffs.propensity)` drug classes used to construct a propensity score.  However a comparison of time windows representing recent and less recent exposure in ever-users did not show a stronger association with recent exposure.

*Conclusions* -- In Scotland severe COVID-19 is strongly associated with polypharmacy. and this association is not easily explained by co-morbidity.  Of all drug classes, proton pump inhibitors make the largest independent contribution to this association.  As prescriptions of this drug class are a recognised indicator for overprescribing of drugs without clear medical indication, it is difficult to distinguish between causality and confounding as explanations for this association.  However the strength of the association in younger age groups and in those not on other drugs with which proton pump inhibitors are widely co-prescribed raise concern about a possible causal effect that requires further investigation.  In the meantime existing advice on reducing inappropriate polypharmacy and specifically limiting the prescribing of proton pump inhibitors should be reinforced.

@@

# Background
The REACT-SCOT study is a matched case control study of severe COVID-19 in Scotland.  In the first report of this study we reported a strong association of severe COVID-19 with having had at least one dispensed prescription in the past year.  The univariate rate ratios associated with at least one dispensed prescription varied from 10 in those aged under 60 years to 40 in those aged 75 years and over.  This association persisted after adjusting for care home residence, any hospital admission in the last five yeasrs and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).  In a multivariate analysis adjusting for these covariates, the rate ratios five years or residence in a care home, the rate ratios associated with any dispensed prescription were 5.0 in those aged under 40 and 11 in those aged 75 years and over.  Partitioning of the "any prescription" variable between BNF chapters, which represent broad indication-based drug classes, showed that prescriptions in almost all BNF chapters were associated with elevated rate ratios for COVID-19. In a multivariate analysis of all chapters, most of these associations were weaker, but prescriptions of at least one drug in chapters 1 (gastrointestinal system), 2 (nervous system) and 9 (nutrition and blood) remained strongly associated with COVID-19.  We concluded that although most associations of severe COVID-19 with drug prescribing were likely to be attributable to confounding by the indications for which these drugs were prescribed or by frailty especially in older persons, causal effects of specific drugs or direct effects of polypharmacy on susceptibility could not be ruled out.

@@

# Background
The REACT-SCOT study is a matched case-control study of severe COVID-19 in Scotland.  In the first report from this study an unexpected result was the strong association of severe COVID-19 with dispensing of at least one prescription in the past year.  The rate ratio associated with this variable varied from 10 in those aged under 60 years to 40
in those aged 75 years and over. The objective of this study was to establish how much of this association could be explained by confounding by co-morbidity and other health-related factors that influece drug prescribing and independently affect susceptibility to severe COVID-19, and how much by direct effects of prescribed drugs.  A secondary objective was to determine how much of the association could be explained by a sparse set of drug classes or other health-related factors, and how much was attributable to multifactorial causes, possibly including interactions between these factors.

# Methods
The design of the REACT-SCOT case-control study is described in detail in the accompanying paper.

## Case ascertainment and sampling of controls 
All individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  For each test-positive case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.  Socioeconomic status was recorded as the quintile of the Scottish Index of Multiple Deprivation, based on the postal code.

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2.

British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.  The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).  BNF subparagraph codes generally represent pharmacological drug classes.  For this analysis prescription codes from chapters 14 and above, comprising mostly dressings and appliances but also vaccines, anaesthesia and other preparations were grouped as "Other".  Defined daily doses (DDDs) of each drug were obtained from the DDD/ATC Index of the WHO Collaborating Centre for Drug Statistics Methodology.  Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to the cutoff of 15 days before the date that the case (in each stratum of one case and ten matched controls) first tested positive.

# Statistical analysis 
Rate ratios for severe COVID-19 were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.  As described previously, we derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  For this analysis the list was extended to include non-cardiac circulatory disease as our analyses of the earlier version of the case-control dataset had shown this to be a risk factor for severe COVID-19.  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that condition is the only indication.  Diagnoses of diabetes were extracted from linkage to the national diabetes register.  The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material.

@@

We constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.  As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, a independent dataset of `r numrows.cchosp` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls. By using this independent dataset to construct the propensity score, we ensured that it would not overfit the case-control dataset.  The propensity score was constructed by stepwise logistic regression with ever-use of a proton pump inhibitor as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription. As the case-control analysis is conditioned on age and sex, the terms for age and sex were dropped when the score was computed in the case-control dataset.

# Results
## Univariate associations of severe COVID-19 with drug classes
We first examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.  Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed for at least 10 individuals and were associated with severe disease at $p < 0.001$.  These COVID-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing without medical indications: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.  This led us to examine associations with polypharmacy, indexed by the number of different drug classes for which prescriptions had been dispensed.

## Relation of severe COVID-19 to number of drug classes dispensed. 
Table \ref{tab:numdrugs.agegr} shows that severe COVID-19 was strongly related to the total number of different drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed since 1 June 2019.  The strength of this association was present in all age groups. With no dispensed prescriptions as reference category, the rate ratios associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.

Table \ref{tab:numdrugs.cats3} shows this relationship after stratifying by care home residence and diagnosis of at least one listed condition.  This shows that the association of severe disease with number of drug classes dispensed was weak in care home residents, absent in those living independently with a diagnosed listed condition, and very strong in those who were living independently without a listed condition. In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.cats3[3, 11]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.cats3[7, 11]` in those with more than 12 drug classes dispensed.  To examine this association further, we restricted the analysis to those not resident in a care home and without listed conditions so as to lessen confounding by underlying disease states and by frailty.

@@

* Examination of the relationship by age.

@@

* Comparison in ever-exposed individuals between recent and less recent time windows of exposure.

@@

### Dose-response relationship
Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inibitor, expressed as average DDDs/day over the period for which prescribing was observed. Table \ref{tab:ppinames} compares the effects of the five drugs in this class.  Omeprazole and lansoprazole were by far the most commonly prescribed drugs.  The effect size, expressed as rate ratio for unit increase in dose expressed as average DDDs/day, was higher for lansoprazole than for omeprazole.  This suggests that the DDDs for lansoprazole (30 mg) and omeprazole (20 mg) are not equivalent in potency.

### Relationship by age
Table \ref{tab:everuse.protonpump} shows the association with ever-use of proton pump inhibitors by age group, without and with adjustment for prespecified covariates.  The association was strongest in those aged under 60 years, and was reduced by adjusting for covariates.  Table \ref{tab:fatal.protonpump} shows the univariate associations of ever-use with fatal COVID-19 by age group.

Table \ref{tab:dosegr.protonpump} shows the rate ratio per unit increase in dose of proton pump inibitor, calculated as average defined daily doses dispensed per day during the observation period.  There is a strong dose-response relationship in those aged less than 75 years, but not in those aged over 75 years.

### Time window comparison
Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.  time window. The rate ratio was not significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.

### Construction of a propensity score
A propensity score for ever-use of a proton pump inhibitor was constructed as described in the Methods section. The `r nrow(coeffs.propensity)` variables retained by stepwise regression, and their coefficients, are shown in Table \ref{tab:propensity}.  As expected, the variables retained included NSAIDs and antiplatelet drugs, but also other drug classes such as tricyclic antidepressants that are not recognized medical indications for co-prescribing of a proton pump inhibitor. Table \ref{tab:bypropensity} shows that when the case-control dataset was stratified by propensity score, the association of COVID-19 with use of proton pump inhibitors remained strong only in those with a propensity score of zero, for whom no prescription had been dispensed for any of the drug classes contributing to the propensity score.  <!--- show coefficient for COVID of each drug class in propensity score before and after adjustment for PPI? --->

## Associations with other drug classes
Tables \ref{tab:bnfchapter1} and \ref{tab:bnfchapter4} show associations with drug classes in BNF chapters 1 (gastrointestinal) and 4 (nervous system).  These chapters were selected as of interest because many drugs in these chapters were associated with COVID-19 even though the indications for which they are prescribed are not conditions recognized as conferring high risk for severe disease.  Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, osmotic laxatives, and aminosalicylates were independently associated with disease in a multivariate analysis.  Proton pump inhibitors were far more commonly prescribed than other drug classes in this chapter.  Table \ref{tab:bnfchapter2} shows that prescriptions of most classes of drugs for the cardiovascular system were associated with elevated rate ratios for severe COVID-19. In a multivariate analysis the strongest associations were with loop diuretics, anticoagulants and antiplatelet drugs.  The univariate rate ratio associated with oral anticoagulants was `r table.bnfchapter2[grep("Oral anticoagulants", rownames(table.bnfchapter2)), 3]`;  Adjustment for other (non-ischaemic) heart disease and ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.anticoagulant.adjusted[1], coeff.anticoagulant.adjusted[3])`.

Tables \ref{tab:bnfchapter2} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 2 (cardiovascular) and 10 (musculoskeletal).  These chapters were selected as of interest because possible effects of drugs in these chapters -- angiotensin-converting enzyme (ACE) inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulation [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.  Of drug classes commonly used to treat hypertension, potassium-sparing diuretics, beta-blockers and angiotensin-converting enzyme inhibitors were associated with increased risk of COVID-19 in univariate analyses, but these associations were weak or absent in a multivariate analysis.  Table \ref{tab:bnfchapter10} shows associations with drugs disaggregated by generic name, as the BNF groups all disease-modifying antirheumatic drugs under a single subparagraph code. All antirheumatic drugs were associated with elevated rate ratios for COVID-19.  The univariate rate ratio associated with hydroxychloroquine sulfate was `r table.bnfchapter10[grep("Hydroxychloroquine", rownames(table.bnfchapter10)), 3]`; adjustment for ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.hydroxychlor.adjusted[1], coeff.hydroxychlor.adjusted[3])`.

# Discussion 
We have shown that severe COVID-19 is associated with dispensed prescriptions of drugs of many different classes.  
These include drugs acting on the the cardiovascular system, where prescribing of multiple drugs is supported by evidence-based guidelines, but also drugs acting on the gastrointestinal and central nervous systems that are widely used as indicators of overprescribing and inappropriate polypharmacy in the National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].  This led us to examine associations of COVID-19 with polypharmacy, defined as the number of drugs classes dispensed during the period of observation.  We attempted to determine how much of this association is attributable to confounding by multimorbidity, and how much is attributable to adverse effects of drug prescribing.  Without more data on health status from primary care records, the confounders cannot be accurately measured.  We were able to show that the association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with common conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.  The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than those associated with designated conditions.  In epidemiological experience rate ratios of this magnitude usually have a causal explanation.

We have shown that a systematic search by backwards elimination identifies proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with number of drug classes dispensed.  The association of COVID-19 with dispensed prescriptions of proton pump inhibitors shows a clear dose-response effect.  Adjustment for covariates pre-specified as likely to confound this assocation -- disorders of the upper gastrointestinal tract, NSAIDs, antiplatelet agents and anticoagulants -- reduces the effect size only slightly. The rate ratio associated with exposure to proton pump inhibitors is highest in those aged under 40, where comorbid conditions and frailty are less likely to confound this association.  The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptiblity to COVID-19: antidepressants, analgesics and nutritional supplements.  When the exposure under study is not measured accurately, including such variables in the model as possible confounders may over-adjust the effect of that exposure under study towards the null.  By restricting the analysis to those with a propensity score of zero, who had no dispensed prescriptions for any of the `r nrow(coeffs.propensity)` drug classes that contribute to the propensity score, we are able to overcome this but the numbers of exposed cases and controls in this category are relatively small.

>>>>

_Background_-- We previously reported that dispensing of prescriptions from primary care in the past year was strongly associated with severe COVID-19 in Scotland, even in those without past diagnoses of conditions designated as high risk (listed conditions). We report a further investigation of possible explanations for this association, which include specific co-morbidity or drug effects but also more diffuse effects of frailty or polypharmacy.

@@

*Findings* -- We began by examining associations with polypharmacy, defined as total number of drug classes dispensed.  The rate ratios (95% CIs) for severe COVID-19 associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.  The association with polypharmacy was strongest in those without past hospital admissions or diagnoses of conditions previously designated as conferring high risk.  The drug classes associated with COVID-19 included widely-prescribed cardiovascular drugs and also multiple drug classes previously identified as indicators of overprescribing without clear medical indications.  A backwards elimination procedure identified proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy.  Further analysis of this association showed a dose-response relationship with the average daily dose dispensed and that the association was strongest in younger age groups: the rate ratio for severe COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[3, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 60 years.  The association remained after adjusting for covariates prespecified as likely confounders, and in those not prescribed any of the `r nrow(coeffs.propensity)` drug classes used to construct a propensity score.  However a comparison of time windows representing recent and less recent exposure in ever-users did not show a stronger association with recent exposure.

*Conclusions* -- In Scotland severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity as it is strongest in those without relevant medical history.  Of all drug classes, proton pump inhibitors make the largest independent contribution to this association.  As prescriptions of this drug class are a recognised indicator for overprescribing of drugs, it is difficult to distinguish between causality and confounding as explanations for this association.  However the strength of the association in younger age groups and in those not on other drugs with which proton pump inhibitors are widely co-prescribed raise concern about a possible causal effect that requires further investigation in other settings.  In the meantime existing advice on reducing inappropriate polypharmacy and limiting the prescribing of proton pump inhibitors should be reinforced.

@@

# Background
The REACT-SCOT study is a matched case control study of severe COVID-19 in Scotland.  In the first report of this study we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year.  The univariate rate ratios associated with at least one prescription varied from 10 in those aged under 60 years to 40 in those aged 75 years and over.  This association persisted after adjusting for care home residence, any hospital admission in the last five yeasrs and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).  Partitioning of the "any prescription" variable between BNF chapters, which represent broad indication-based drug classes, showed that prescriptions in almost all BNF chapters were associated with elevated rate ratios for COVID-19.  We concluded that although most associations of severe COVID-19 with drug prescribing were likely to be attributable to confounding by the indications for which these drugs were prescribed or by frailty especially in older persons, causal effects of drugs on disease susceptibility could not be ruled out.

@@

# Methods
The design of the REACT-SCOT case-control study is described in detail in the accompanying paper. All individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  For each test-positive case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.  Socioeconomic status was recorded as the quintile of the Scottish Index of Multiple Deprivation (SIMD), based on the postal code.

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2.  British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.  The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).  BNF subparagraph codes generally represent pharmacological drug classes.  For this analysis prescription codes from chapters 14 and above, comprising mostly dressings and appliances but also vaccines, anaesthesia and other preparations were grouped as "Other".  Defined daily doses (DDDs) of each drug were obtained from the DDD/ATC Index of the WHO Collaborating Centre for Drug Statistics Methodology.  Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to the cutoff of 15 days before the date that the case (in each stratum of one case and ten matched controls) first tested positive.

# Statistical analysis 
Rate ratios for severe COVID-19 were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.  As described previously, we derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  For this analysis the list was extended to include non-cardiac circulatory disease as analysis of the earlier version of the case-control dataset had shown this to be a risk factor for severe COVID-19.  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an dispensed prescription of a drug for which that condition is the only indication.  Diagnoses of diabetes were extracted from linkage to the national diabetes register.  The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material.

@@

To study the effect of a specific drug, we constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.  As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, a independent dataset of `r numrows.cchosp` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls. By using this independent dataset to construct the propensity score, we ensured that it would not overfit the case-control dataset.  The propensity score was constructed by stepwise logistic regression with ever-use of the drug under study as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription. `r nrow(coeffs.propensity)` drug classes were retained in the propensity score.  As the case-control analysis is conditioned on age and sex, the terms for age and sex were dropped when the score was computed in the case-control dataset.

# Results
## Relation of severe COVID-19 to number of drug classes dispensed. 
We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive). Table \ref{tab:numdrugsgr.agegr} shows that risk of severe COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.

Table \ref{tab:numdrugs.num.icdch} shows that this association was strongest in those who had no hospital diagnoses (because they had not been admitted to hospital).  In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.num.icdch[2, 4]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.num.icdch[6, 4]` in those with more than 12 drug classes dispensed. Table \ref{tab:numdrugs.listed.any} shows that this association was strongest in those without a diagnosis of a listed condition.

To study the relationship further, we restricted the analysis to those not resident in a care home and without a diagnosis of a listed condition.  The rationale for this restriction was to reduce the effect of factors including multi-morbidity and frailty which might be expected to confound the association between polypharmacy and severe COVID-19.  In people with listed conditions, we might expect prescribing of multiple drugs in accordance with evidence-based guidelines for management of these conditions.  Residents in care homes are at high risk of severe COVID-19 for reasons unrelated to polypharmacy, and the profile of drugs most commonly prescribed to people in care homes differs from the profile of drugs most commonly prescribed to people living independently.

## Univariate associations of severe COVID-19 with drug classes
We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.  Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.  These COVID-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing without medical indications: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.

@@

@@

* Comparison between recent and less recent time windows of exposure.

@@

Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inhibitor, expressed as average DDDs/day over the period for which prescribing was observed. This relationship is strongest in younger age groups, and the effect size is reduced by about one quarter by adjusting for prespecified covariates. 
Of the five drugs in this class in the BNF, only two -- omeprazole and lansoprazole -- were widely prescribed: mean DDDs/day in controls were `r table.ppinames[3, 1]` and `r table.ppinames[2, 1]` respectively.  The effect size for unit increase in dose expressed as average DDDs/day was higher for lansoprazole [rate ratio `r table.ppinames[2, 3]`] than for omeprazole rate [rate ratio `r table.ppinames[3, 3]`].  This suggests that the DDDs for lansoprazole (30 mg) and omeprazole (20 mg) are not equivalent in potency.

Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.  The rate ratio was not significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.

A propensity score for ever-use of a proton pump inhibitor was constructed as described in the Methods section. The `r nrow(coeffs.propensity)` drug classes retained by stepwise regression, and their coefficients, are shown in Table \ref{tab:propensity}.  As expected, the drug classes retained included NSAIDs and antiplatelet drugs, but also other drug classes such as tricyclic antidepressants that are not recognized medical indications for co-prescribing of a proton pump inhibitor. Table \ref{tab:bypropensity} shows that when the case-control dataset was stratified by propensity score, a strong association of COVID-19 with use of proton pump inhibitors remained in those with a propensity score of zero, for whom no prescription had been dispensed for any of the drug classes contributing to the propensity score.

## Associations with other drug classes
Tables \ref{tab:bnfchapter1}, \ref{tab:bnfchapter2}, \ref{tab:bnfchapter4} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 1 (gastrointestinal) 2 (cardiovascular), 4 (nervous system) and 10 (musculoskeletal).  These chapters were selected as of interest because many drugs in these chapters were associated with COVID-19 or because hypotheses about possible effects of drugs in these chapters -- angiotensin-converting enzyme (ACE) inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.

Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, osmotic laxatives, and aminosalicylates were independently associated with disease in a multivariate analysis.  Proton pump inhibitors were far more commonly prescribed than other drug classes in this chapter.

Table \ref{tab:bnfchapter2} shows that prescriptions of most classes of drugs for the cardiovascular system were associated with elevated rate ratios for severe COVID-19. In a multivariate analysis the strongest associations were with loop diuretics, anticoagulants and antiplatelet drugs.  Of drug classes commonly used to treat hypertension, potassium-sparing diuretics, beta-blockers and angiotensin-converting enzyme inhibitors were associated with increased risk of COVID-19 in univariate analyses, but these associations were weak or absent in a multivariate analysis. The univariate rate ratio associated with oral anticoagulants was `r table.bnfchapter2[grep("Oral anticoagulants", rownames(table.bnfchapter2)), 3]`; Adjustment for other (non-ischaemic) heart disease and ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.anticoagulant.adjusted[1], coeff.anticoagulant.adjusted[3])`.

Table \ref{tab:bnfchapter4} shows that the strongest associations of COVID-19 with drugs for the nervous system were with drugs widely used in care homes: antipsychotics, analgesics and drugs for dementia.

Table \ref{tab:bnfchapter10} shows associations with drugs disaggregated by generic name, as the BNF groups all disease-modifying antirheumatic drugs under a single subparagraph code. All antirheumatic drugs were associated with elevated rate ratios for COVID-19.  The univariate rate ratio associated with hydroxychloroquine sulfate was `r table.bnfchapter10[grep("Hydroxychloroquine", rownames(table.bnfchapter10)), 3]`; adjustment for ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.hydroxychlor.adjusted[1], coeff.hydroxychlor.adjusted[3])`.

# Discussion 
We have shown that severe COVID-19 is associated with dispensed prescriptions of drugs of many different classes.  These include drugs acting on the the cardiovascular system, where prescribing of multiple drugs is supported by evidence-based guidelines, but also drugs acting on the gastrointestinal and central nervous systems that are widely used as indicators of overprescribing and inappropriate polypharmacy in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].  This led us to examine associations of COVID-19 with polypharmacy, defined as the number of drugs classes dispensed during the period of observation.  We attempted to determine how much of this association is attributable to confounding by multimorbidity, and how much is attributable to adverse effects of drug prescribing.  Without more data on health status from primary care records, the confounders cannot be accurately measured.  We were able to show that the association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with common conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.  The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the risk ratios of about 3 for severe COVID-19 that are associated with designated conditions.

We have shown that a systematic search by backwards elimination identifies proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with number of drug classes dispensed.  The association of COVID-19 with dispensed prescriptions of proton pump inhibitors shows a clear dose-response effect.  Adjustment for covariates pre-specified as likely to confound this assocation -- disorders of the upper gastrointestinal tract, NSAIDs, antiplatelet agents and anticoagulants -- reduces the effect size only slightly. The rate ratios associated with exposure to proton pump inhibitors are highest in younger age groups, in whom comorbid conditions and frailty are less likely to confound this association.  The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptiblity to COVID-19: antidepressants, analgesics and nutritional supplements.  When the exposure under study is not measured accurately, including such variables in the model as possible confounders may over-adjust the effect of that exposure under study towards the null.  By restricting the analysis to those with a propensity score of zero, who had no dispensed prescriptions for any of the `r nrow(coeffs.propensity)` drug classes that contribute to the propensity score, we are able to overcome this but the numbers of exposed cases and controls in this category are relatively small. One of the most powerful pharmacoepidemiologic methods for distinguishing causality from confounding is to investigate the relationship of adverse events to the time of exposure among ever-users of the drug under study.  We did not detect any difference between the rate ratios associated with dispensing of a proton pump inhibitor in more recent and less recent time windows.  However most users of these drugs had prescriptions dispensed in both time windows.  Users whose most recent prescription was in the earlier time window may have had a stock of tablets for use as required.  It is also possible that use of proton pump inhibitors has a persistent effect on susceptibility to infection through alteration of the gut microbiome.
